\setdictum{%
  Who are you? How did you get in my house?%
}{%
  Donald E. Knuth about one-based array indices in algorithms
  (according to \texttt{xkcd}\footnotemark)%
}

\longchapter{%
  Algorithms for B-Splines on Sparse Grids%
}{%
  Algorithms for B-Splines on\texorpdfstring{\\}{ }Sparse Grids%
}{%
  Algorithms for B-Splines on Sparse Grids%
}
\footnotetext{\url{https://xkcd.com/163/}}
\label{chap:40algorithms}

\initial{0.6em}{L}{ittle is known}
about the algorithmic challenges
that hierarchical bases of B-spline type (or even of general type) pose
on sparse grids.
In general, we are not able to directly apply the sparse grid algorithms
that were designed for hat functions $\bspl{l,i}{1}$.
Hence, we have to generalize these algorithms
to higher-order B-splines $\bspl{l,i}{p}$
or even to arbitrary tensor product basis functions.
The main problem is the larger support
of higher-order B-splines when compared to degree $p = 1$.
The larger support introduces new dependencies between
values of grid points that cannot be resolved with conventional algorithms.

This chapter gives an overview of six algorithms for B-splines
on sparse grids.
Two of these algorithms are already known
\multicite{Griebel92Combination,Balder94Adaptive},
while the remaining four are new.
Correctness results are given for every algorithm.
We use hierarchization as the exemplary problem for our algorithms,
but the ideas of the algorithms can be generalized to any linear operator.
Furthermore, most algorithms are not tailored to B-splines $\bspl{l,i}{p}$,
but applicable to general tensor product basis functions $\basis{l,i}$.
Whether an algorithmic approach is feasible for the sparse grid at hand
or not depends on the grid's type:
full grid, dimensionally adaptive sparse grid, or
spatially adaptive sparse grid.
The more assumptions the grid satisfies, the faster and
easier the corresponding algorithms will be.

\Cref{sec:41problem} explains hierarchization as our example problem
and defines the notation used in this chapter.
The remaining four sections treat the three different types of grids:
First, \cref{sec:42fullGrids} deals with full grids to formalize and repeat
the well-known \up.
Second, \cref{sec:43dimAdaptive} focuses on algorithms for
dimensionally adaptive sparse grids.
Third, \cref{sec:44spatAdaptiveBFS} and \cref{sec:45spatAdaptiveUP}
treat arbitrary (spatially adaptive) sparse grids,
which is the most interesting case for us.
\Cref{sec:44spatAdaptiveBFS} employs \bfs for hierarchization,
while \cref{sec:45spatAdaptiveUP} uses the \up\punctfix{.}

This original chapter is the main theoretical contribution of this thesis.
Although the unidirectional principle in \cref{sec:42fullGrids} and
the combination technique in \cref{sec:43dimAdaptive} are well-known,
the presentation with formal proofs of correctness is new.
Parts of the chapter have already been published in scientific papers,
namely \cref{sec:44spatAdaptiveBFS} \cite{Valentin18Fundamental}.
The weakly fundamental splines (\cref{sec:454wfs}) and the
Hermite hierarchization method (\cref{sec:455hermiteHierarchization})
are based on an idea by Dr.\ Stefan Zimmer (University of Stuttgart, Germany).



\input{tex/document/41problem}
\input{tex/document/42fullGrids}
\input{tex/document/43dimAdaptive}
\input{tex/document/44spatAdaptiveBFS}
\input{tex/document/45spatAdaptiveUP}

\cleardoublepage
