\setdictum{%
  Money. A social life. A shave.\\
  A Ph.D.\ student needs not such things.%
}{%
  Mike Slackenerny
  (PHD Comics\footnotemark)%
}

\longchapter{%
  Application 1: Topology Optimization%
}{%
  Application 1:\texorpdfstring{\\}{ }Topology Optimization%
}{%
  Application 1 -- Topology Optimization%
}
\footnotetext{\url{http://phdcomics.com/comics/archive.php?comicid=40}}
\label{chap:60topoOpt}

\initial{0em}{N}{ow, we want to investigate}
the first real-world application,
which is the field of topology optimization.
The classical and widely-used method in engineering is shape optimization,
where the shape of a component
(parameterized by $\*x \in \real^d$) has to be
determined such that some objective function value $\objfun(\*x)$ is optimal,
i.e., minimal or maximal.
For example, a bridge over a valley can be built in the shape of a
parabolic arc.
The task of shape optimization is then to choose the coefficients of the
parabola such that the bridge's stability is maximized,
possibly with the constraint that the volume occupied by the bridge
does not exceed a certain value (to save construction costs) or
that the size of the resulting passage meets some size requirements
(e.g., at least \SI{20}{\meter} wide and \SI{6}{\meter} tall).

However, the framework of shape optimization unnecessarily prescribes the
topology of the shapes in the search space \cite{Allaire16Towards}.
In the bridge example, it may well be that a viaduct-type bridge with
three arcs instead of one is more stable while occupying less volume.
We are not able to find such a bridge with shape optimization
in the previous example,
as we have restricted the search space to single-arc bridges.
This issue is resolved by the more sophisticated
framework of topology optimization, where the topology%
\footnote{%
  Two objects are considered ``topologically different''
  if their numbers of ``holes'' differ.
  This stems from the fact that in the field of mathematical topology,
  the \term{genus} (i.e., the number of holes)
  of a topological space is a \term{topological invariant,} i.e.,
  the genus is invariant under homeomorphism.
  If the genera of two topological spaces differ, then they cannot be
  homeomorphic and are thus considered topologically different.%
}
is not given by the user,
but chosen by the optimization algorithm (in a hopefully optimal way),
rendering topology optimization a key area of simulation technology.

Recently, B-splines have been used for
shape optimization \cite{Martin16Formoptimierung} and
topology optimization \multicite{Qian13Topology,Zhang17Topology}.
Sparse grids have been employed for
topology optimization \cite{Huebner14Mehrdimensionale} as well.
In this chapter, we want to combine these two numerical tools,
which have been used in isolation until now,
to perform topology optimization using B-splines on sparse grids.
The two most common approaches for topology optimization are
the \term{level-set method} and
the \term{homogenization method} \cite{Allaire16Towards}.
The level-set method describes the boundary of the object
as the zero level set $\psi^{-1}(0)$ of a function
$\psi\colon \objdomain \to \real$ \term{(level-set function)}
and uses a \pde to iteratively transport this function and,
consequently, the object's boundary \cite{Allaire04Topology}.
However, we want to focus on the second method:
the method of homogenization.

This chapter is structured as follows:
\Cref{sec:61homogenization} explains the homogenization method.
In \cref{sec:62tensors}, we discuss the details of applying B-splines on
sparse grids to this method.
We set up different micro-cell models and scenarios in \cref{sec:63models},
before reviewing numerical results in \cref{sec:64results}.
The results in this chapter have been obtained in collaboration with
Prof.\ Dr.\ Michael Stingl and Daniel Hübner
(both FAU Erlangen-Nürnberg, Germany).
The author of this thesis contributed the parts related to
interpolation and sparse grids, while the collaborators at FAU
studied the engineering and application parts of the joint project
(for example, they provided optimization scenarios and
assessed the quality of the results).

\input{tex/document/61homogenization}
\input{tex/document/62tensors}
\input{tex/document/63models}
\input{tex/document/64results}

\cleardoublepage
