\setdictum{%
  Beware of bugs in the above code;\\
  I have only proved it correct, not tried it.%
}{%
  Donald E.\ Knuth \cite{Knuth77Notes}%
}

\longchapter{%
  Application 2: Musculoskeletal Models%
}{%
  Application 2:\texorpdfstring{\\}{ }Musculoskeletal Models%
}{%
  Application 2 -- Musculoskeletal Models%
}
\label{chap:70muscle}

\initial{0em}{E}{xisting musculoskeletal models of muscle-tendon complexes,}
e.g., of the human upper limb, can mainly be divided into two different types.
\term{Lumped-parameter musculoskeletal models,}
for example Hill-type models based on multi-body simulations
\multicite{Roehrle16Two,Valentin18Gradient},
constitute the most common type.
These models assume that the components of the musculoskeletal system
are rigid.
The mechanics is reduced to point masses
associated with their moment of inertia;
thus, these models can be described by few parameters.

\term{Continuum-mechanical musculoskeletal models} form the other type.
Their advantage is that they are more detailed and, hence, more realistic.
However, their increased complexity leads to higher computational costs.
As an example, we consider an inverse problem (see \cref{chap:10introduction})
that involves a continuum-mechanical simulation of such a
musculoskeletal model,
where we search values of model parameters
such that a specific movement is attained.
Each iteration of the solution process for such an inverse problem
may take hours or even days, depending on the model at hand.

Surrogate methods based on sparse grids help to decrease the complexity
in two ways:
First, the evaluation of surrogates is obviously drastically cheaper
than the solution of the inverse problem.
Second, the particular choice of sparse grids decreases the number
of necessary samples to construct the surrogates.
As for the previous application,
the choice of B-splines as hierarchical basis functions enables
the evaluation of continuously differentiable surrogate gradients.
For the example of inverse problems, this means that
gradient-based optimization methods may be employed,
which significantly accelerates convergence compared to gradient-free methods.

This chapter is split into three sections.
First, in \cref{sec:71model}, we introduce a continuum-mechanical
model of the human upper limb.
Second, in \cref{sec:72methodology}, we list the types of inverse problems
of interest.
Finally, in \cref{sec:73results}, we present numerical results
regarding the solution of these inverse problems.

The results of this chapter are based on a collaboration with
Prof.\ Oliver Röhrle, PhD, and Dr.\ Michael Sprenger
(both SimTech/University of Stuttgart, Germany).%
\footnote{%
  Michael Sprenger left the University of Stuttgart in 2015.%
}
The collaborators contributed the biomechanical model
with its theory, its geometry, and its implementation,
while the author of this thesis contributed the
sparse grid/B-spline methodology and
computed the numerical results.
Note that the results have already been
published in a paper \cite{Valentin18Gradient},
which we will follow closely in this chapter.

\input{document/71model}
\input{document/72methodology}
\input{document/73results}

\cleardoublepage
