\setdictum{%
  The goal is to buy as many iPads as possible during your lifetime.%
}{%
  In a talk at the 5th Workshop on\\Sparse Grids and Applications%
}

\longchapter{%
  Application 3: Dynamic Portfolio Choice Models%
}{%
  Application 3:\texorpdfstring{\\}{ }Dynamic Portfolio Choice Models%
}{%
  Application 3 -- Dynamic Portfolio Choice Models%
}
\label{chap:80finance}

\initial[lhang=0.06]{0.1em}{S}{urrogates based on B-splines on sparse grids}
can also be used for our third application,
which stems from finance.
In this application, we optimize financial decisions of an individual
over their lifetime in discrete time steps or iterations $t = 0, \dotsc, T$
(for example, years $t = 0, \dotsc, 80$, where $20+t$ is the age
of the individual), depending on internal and external factors.
There are three types of variables:

\begin{itemize}
  \item
  \term{State variables} $\state_t$
  such as the individual's wealth $\wealth_t$ and their income
  cannot be controlled directly by the individual.
  Instead, the individual's decisions may influence the value of
  state variables of future iterations.%
  \footnote{%
    The time $t$ can also be regarded as a state variable.%
  }
  
  \item
  \term{Policy variables} $\policy_t$
  such as consumption $\consume_t$ and the amount of stocks to buy or sell
  represent the investment decisions the individual can make in
  each iteration.
  They are subject to specific constraints
  (for instance, you cannot spend more money than you have,
  if you do not allow debts).
  
  \item
  \term{Stochastic variables} $\stochastic_t$
  such as return rates of stocks and inflation
  cannot be controlled by the individual at all.
  Therefore, statements about optimal investment conditions
  are usually made for expected values instead of exact values.
\end{itemize}

\noindent
We discretize the state space with a spatially adaptive sparse grid.
For each state grid point, an optimization problem over the policy
variables has to be solved, where the objective function depends
on the expected value over the stochastic variables.
By using B-splines as hierarchical basis functions,
the accuracy of the interpolants is increased and
the explicitly known gradients enable the usage of
gradient-based optimization methods, thus accelerating convergence.
The process is repeated for each time step,
which is possible due to the Bellman principle,
which implies that the objective functions
occurring at time $t$ depend on the interpolant of the next iteration $t+1$.
Hence, the problem has to be solved backward in time
via a scheme that closely resembles dynamic programming.

The outline of this chapter is as follows:
In \cref{sec:81models}, we formalize the framework of
dynamic portfolio choice models and describe our approach.
Afterwards, we explain in \cref{sec:82algorithms} the necessary algorithms
for implementing the solution of these models.
\Cref{sec:83problem} introduces the transaction costs problem
as an example application of the general framework presented
in \cref{sec:81models}.
Finally, in \cref{sec:84results}, we study numerical results.

This chapter is based on a collaboration with Prof.\ Dr.\ Raimond Maurer
and Peter Schober (both Goethe University Frankfurt, Germany).
In previous work, Peter Schober treated the solution of
dynamic portfolio choice models with piecewise linear basis functions
on spatially adaptive sparse grids \cite{Schober18Solving}.
The original contribution of this thesis is the introduction
of higher-order B-splines for the solution of these problems.
The author of this thesis contributed the methodology of
hierarchical B-splines and large parts of the implementation.
The contributions of the collaborators at Goethe University Frankfurt
are the financial models, the literature review of related work,
and the assessment of the quality of results.

\input{document/81models}
\input{document/82algorithms}
\input{document/83problem}
\input{document/84results}

\cleardoublepage
